### 2016-05-25 Paper_v0.tex
Merged changes to intro refs results and figs

\section*{Introduction}
Complement is a central part of innate immunity and plays a very significant role in regulating the inflammatory response. It was first discovered in the 1890s where it was found to 'complement' the bactericidal activity of natural antibodies. Complement is mediated through a set of approximately 30-35 soluble and cell surface proteases \cite{}. The central process in complement activation involves the formation of Membrane Attack Complex (MAC) and a protein called C5a. Complement activation takes places through three different pathways: the alternate, the classical and the lectin. Each of these pathways involves a different initiator signal that leads to the formation of a serine protease called C5 convertase which cleaves an inactive protein called C5 to form C5a and C5b. The classical pathway is triggered when antibodies form complexes with foreign antigens or other pathogens. A multimeric protein complex C1 binds to the antigen-antibody complex and undergoes a conformational change. This activated complex cleaves proteins C4 and C2 to C4a, C4b, C2a and C2b respectively. C4a and C2b combine to form a protease C4bC2a also known as the classical C3 convertase. The lectin pathway is initiated through the binding of L-ficolin or Mannose Binding Lectin (MBL) to the carbohydrates on the surfaces of bacterial pathogens. This bound complex in turn cleaves C4 and C2 and leads to the production of C4bC2a. The alternate pathway involves a 'tickover' mechanism in which a protein called C3 is hydrolyzed to form C3b. In presence of foreign pathogens C3b binds to these surfaces and recruits  additional factors called factor B and factor D that lead to the formation of alternate C3 convertase - C3bBb. The formation of classical and alternate C3 convertases on bacterial surfaces is followed by the formation of proteases called C5 convertases. The classical and alternate C3 convertases recruit C3, Factor B and Factor D to form classical C5 convertase (C4bC2aC3b) and alternate C5 convertase (C3bBbc3B) respectively. The C5 convertases then cleave C5 to form C5a and C5b respectively. 	The cleavage of C5 is followed by a series of sequential cleavages of proteins C6, C7, C8 and C9 that combine with C5b to form the MAC complex. The activation of complement and formation of C5a and MAC complex is regulated at different points through a number of plasma and host cell proteins. The initiation of the classical pathway through the attachment of C1 to an antibody is controlled by the C1 Inhibitor (C1-Inh), a protease inhibitor belonging to the serpin superfamily. C1-Inh irreversibly binds to and deactivates the active subunits of component C1 to prevent spontaneous fluid phase and chronic activation of complement \cite{walker1995complement}. The serum and host-tissue regulation of the upstream elements of the complement system is also achieved through the binding of C4 binding protein (C4BP) to C4b and through the binding of factor H to C3b \cite{blom2001structural}. These proteins are also capable of binding their respective components in the convertase form. Membrane cofactor protein (MCP or CD46) possesses a cofactor activity for C4b and C3b, which protects the host from self-activation of complement \cite{riley2004cd46}.  Decay accelerating factor (DAF or CD55) is able to recognize and dissociate both convertases \cite{lukacik2004complement}. MAC is inhibited by vitronectin and clusterin in the plasma and CD59 at the host surface.
Proteins C3a, C4a, and C5a are inactivated or reduced in activity by carboxypeptidase-N \cite{liszewski1995control}. Research over the past decade has also shown that the role of complement extends beyond as a simple player in the immune system. It is an important network in maintaining homeostasis and malfunction of complement has been linked with several diseases like Alzheimers, acute macular degeneration (AMD), glaucoma, Parkinson's disease, multiple sclerosis, renal failure, schizophrenia, rheumatoid arthritis, cancer, sepsis and coagulopathy. The ability to maintain cross talk with other networks like coagulation, autonomous nervous response and the ability to regulate inflammation makes it important to understand complement in a more holistic perspective.

Thus given the complexity and importance of complement in influencing the human condition, developing models of complement within an integrative framework (that includes other biochemical networks) are crucial to understanding its role in human disease. Traditionally, complement models have been formulated as linear or non-linear Ordinary Differential Equation (ODE) systems. Hirayama  et al. (ref) used a system of linear ODEs to model the classical pathway of complement. Korotaevskiy and co-workers (ref) built a theoretical model of complement using a system of non-linear ODEs that included classical, lectin and alternate pathways. However both these studies involve no validation studies with experimental data. Liu et al analyzed the formation of classical and lectin C3 convertases and the regulatory role of C4BP using a system of 45 non-linear ODEs with 85 parameters.  Recently, Zewde and co-workers built a detailed mechanistic model of alternative complement activation was built using 107 ODEs and 74 kinetic parameters (Ref). This model delineated the response of complement on a host cell and a foreign antigen. However, these previous models were largely based upon mechanistic knowledge.  Given the complexity of complement and its interactions with other networks it is unfeasible and computationally expensive to build such large mechanistic models. In addition is much more difficult to experimentally interrogate the response of various complement proteins under different conditions. This also presents with the problem of estimation of a large number of parameters with little or no experimental data. Thus there exists a need to reduce the mechanistic complexity while capturing dynamics of complement accurately.

@@

We tested the predictive power of the reduced order human complement model with validation data sets not used during model training. Six validation data sets were used, three for C3a and C5a respectively at different zymosan concentrations. All kinetic and control parameters were
fixed for the validation simulations. The reduced order model predicted the C3a and C5a time-course profiles at a qualitative level (Fig. \ref{fig-prediction}). [AND THEN WHAT??]

@@

Though the role of complement in immune response has been well known since long, there has been a paucity of mathematical models of complement. To our knowledge this is the first model of complement that combines different pathways of initiation and validates the dynamics of downstream proteins like C5a using experimental data. Liu and co-workers modeled formation of C3a through the classical pathway using 45 non-linear ODEs  \cite{liu2011computational}. The hybrid modeling framework, however, allowed us to model the lectin mediated formation of C3a using only 5 ODEs.  Though we do not model all the interactions of initiation in detail, especially the cross-talk between lectin and classical pathways like Liu et al. \cite{liu2011computational} we successfully captured C3a dynamics with respect to different concentrations of lectin initiators. The model was also surprisingly accurate in capturing the quantitative dynamics of C3a and C5a formed from the alternate pathway with only 7 ODEs. The lag phase in the initiation of C5a is followed by an acceleration in the production of C5a. This is also qualitatively similar to the predicted C5a time profiles using a theoretical model of complement with 107 equations \cite{zewde2016quantitative}. Similarly, our model was able to capture C3a formation from the alternate pathway that showed the same qualitative trends as Zewde et al \cite{zewde2016quantitative}. We also observe in our model simulation that the quantity of C3a produced in the alternate pathway is nearly 1000 times the quantity of  C5a produced. Though this is in agreement with the experimental data  \cite{morad2015time}, it differs from the theoretical predictions by Zewde et al.\cite{zewde2016quantitative} who show that C3a is ${10^8}$ times the C5a concentration. It is possible that the experimental system has trace amounts of C5a which leads to this difference. In our model the time profile of C5a generation from the lectin pathway changes with respect to the quantity of zymosan (the lectin pathway initiator). We see that the lag phase for generation progressively decreases with increasing concentration of the initiator. Korotaevskiy et al. \cite{korotaevskiy2009non} show a similar lag phase followed by an accelerated production of C5a using a theoretical model of complement for much smaller time scales. Zewde et al. \cite{zewde2016quantitative} show a similar time profile for C5a generated in the alternate pathway as well. We do observe a similar trend in our model. However given the different time scale in our study we do not observe a very prominent lag phase. Our model also captures the maximum concentrations of C5a with different initiators and in the alternate pathway. However we do not capture the maximum concentration of C3a with low initiator levels though we capture the overall trend which includes the initiation and amplification phases. One of the possible reasons for this could be the non-inclusion of C2-by pass pathway which further accelerates production of C3a without the involvement of a C3 convertase. Currently the C3a in the model is generated only through the activity of a C3 convertase. Incorporating this additional step within the reduced order modeling framework would be a future direction that we need to consider. Taken together we surprisingly do very well in capturing the dynamics of key complement proteins and observe similar trends as in large mechanistic models.

%%% We talk about complement therapeutics here in relation with sensitivity analysis

The global sensitivity analysis of our reduced order model provides interesting insights into the regulation of C3a and C5a which are important therapeutic targets.  We see that the formation of C5a in the absence of an initiator is the most sensitive to the formation of basal C3b through the tick over mechanism. This mechanism could be a potential target for C5a inhibitors especially in cases of autoimmune disorders where there is an absence of initiators of the lectin or the classical pathway. In presence of an initiator we see that C5a formation is primarily sensitive to the lectin initiation parameters and the kinetic parameters that govern the conversion of C5 to C5a and C5b. This result agrees well with the current protease inhibitors targeting initiating complexes that includes mannose-associated serine proteases 1 and  2 (MASP-1,2). The most commonly used anti-complement drug eculizumab, targets C5 protein which is cleaved to form C5a. Our sensitivity analysis shows that kinetic parameters governing C5 conversion are sensitive in both lectin initiated and alternate pathways, thus agreeing with targeting C5 protein. The formation of basal C3b is also a sensitive parameter in the formation of C3a through the alternate pathway. Thus this mechanism can act as a target for both C3a and C5a inhibitors. The lectin initiated formation of C3a shows a number of sensitive parameters. This includes the lectin initiation parameters that control the formation of C5a along with C3 convertase inhibition using C4BP and the parameter governing C3 convertase. All these mechanisms involve potential drug targets and literature shows that C4BP, C3 convertase inhibitors are the being developed as the next generation inhibitors of complement \cite{morgan2015complement}. Taken together it is surprising to see that reduced order model has the ability to identify mechanisms for potential drug targets that are in agreement with existing literature.
%Though it is known complement malfunction causes a number of diseases \cite{ricklin2007complement} developing complement specific therapeutics has been challenging\cite{ricklin2007complement}. Currently there are only four approved drugs targeting complement \cite{morgan2015complement}. Eculizumab targets C5, Berinert, Ruconest and Cinryze target C1.

@@

\subsection*{Global sensitivity analysis of model performance}
We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{SOBOL_METHOD}. We computed the total sensitivity index of each parameter relative to four performance objectives, each objective was based on the sum of squared errors between model and experimental data for C3a alternate, C5a alternate, C3a lectin, and C5a lectin simulations. We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble.
We used the sampling method of Saltelli \textit{et al.} \citep{Saltelli:2010} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges,
where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 200 and $d$ = 28, so the total sensitivity indices were computed from
11,600 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.

@@

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure4_Sensitivity_Analysis_may24.pdf}
\caption{Sobol's sensitivity analysis of the reduced order complement model with respect to the modeling parameters.  Sensitivity analysis was conducted on the four cases we used to train our model: (A) C3a at 0 zymosan, (B) C5a 0 zymosan, (C) C3a 1 g zymosan, and (D) C5a $1$ g zymosan. The bars denote total sensitivity index which includes local contribution of each parameter and global sensitivity of significant pairwise interactions. The error bars are the 95 percent confidence interval. $k$ represents association rate, $km$ denote Michaelis-Menten saturation constants, and $alpha$ and $n$ refers to the exponentials of the control functions.}\label{fig-SA}
\end{figure}

>>>>

\section*{Introduction}
Complement is a central part of innate immunity and plays a very significant role in regulating the inflammatory response. Complement was first discovered in the 1890s where it was found to 'complement' the bactericidal activity of natural antibodies. Complement is mediated through a set of approximately 30-35 soluble and cell surface proteases. The central process in complement activation involves the formation of Membrane Attack Complex (MAC) and a protein called C5a. C5a acts as a bridge between innate and adaptive immunity and plays a very important role in regulating inflammation and coagulation.  Complement activation takes places through three different pathways: the alternate, the classical and the lectin. Each of these pathways involves a different initiator signal that leads to a cascade of downstream reactions in the complement system. The classical pathway is triggered when antibodies form complexes with foreign antigens or other pathogens. A multimeric protein complex C1 binds to the antigen-antibody complex and undergoes a conformational change. This activated complex cleaves proteins C4 and C2 to C4a, C4b, C2a and C2b respectively. C4a and C2b combine to form a protease C4bC2a also known as the classical C3 convertase. The lectin pathway is initiated through the binding of L-ficolin or Mannose Binding Lectin (MBL) to the carbohydrates on the surfaces of bacterial pathogens. This bound complex in turn cleaves C4 and C2 and leads to the formation of C4bC2a. The alternate pathway involves a 'tickover' mechanism in which  C3 is hydrolyzed to form C3b. In presence of foreign pathogens C3b binds to these surfaces and recruits  additional factors called factor B and factor D that lead to the formation of alternate C3 convertase - C3bBb. The formation of classical and alternate C3 convertases on bacterial surfaces is followed by the formation of proteases called C5 convertases. The classical and alternate C3 convertases recruit C3, Factor B and Factor D to form classical C5 convertase (C4bC2aC3b) and alternate C5 convertase (C3bBbc3B) respectively. The C5 convertases then cleave C5 to form C5a and C5b respectively. The cleavage of C5 is followed by a series of sequential cleavages of proteins C6, C7, C8 and C9 that combine with C5b to form the membrane attack complex (MAC).
The activation of complement and formation of C5a and MAC  is regulated at different points through a number of plasma and host cell proteins.
The initiation of the classical pathway through the attachment of C1 to an antibody is controlled by the C1 Inhibitor (C1-Inh), a protease inhibitor belonging to the serpin superfamily. C1-Inh irreversibly binds to and deactivates the active subunits of component C1 to prevent spontaneous fluid phase and chronic activation of complement \cite{walker1995complement}. The serum and host-tissue regulation of the upstream elements of the complement system is also achieved through the binding of C4 binding protein (C4BP) to C4b and through the binding of factor H to C3b \cite{blom2001structural}. These proteins are also capable of binding their respective components in the convertase form. Membrane cofactor protein (MCP or CD46) possesses a cofactor activity for C4b and C3b, which protects the host from self-activation of complement \cite{riley2004cd46}.  Delay accelerating factor (DAF or CD55) is able to recognize and dissociate both convertases \cite{lukacik2004complement}. MAC is inhibited by vitronectin and clusterin in the plasma and CD59 at the host surface.
Carboxypeptidase-N, a regulator of inflammation is a plasma zinc metalloprotease that cleaves carboxyl-terminal arginines and lysines of complement complement components and  anaphylatoxins such as C3a, C4a, and C5a.
\cite{liszewski1995control}.

Research over the past decade has also shown that the role of complement extends beyond as a simple player in the immune system. Complement has been shown to contribute to homeostasis by inducing growth factors involved in tissue repair \cite{ricklin2010complement}. The malfunction of complement has been linked with several diseases including Alzheimers, glaucoma, Parkinson's disease, multiple sclerosis, schizophrenia, rheumatoid arthritis and sepsis \cite{ricklin2007complement, rittirsch2008harmful}. Complement has also been shown to play both a positive and negative role in certain cancers; attacking tumor cells with altered surface patterns in some cases potentially contributing to tumor growth in others \cite{sarma2011complement, ricklin2013complement}. Several networks maintain cross talk with complement including coagulation, autonomous nervous response and the ability to regulate inflammation \cite{ricklin2013complement}. The system's involvement in a variety of beneficial regulatory processes coupled with the role it can play in disease makes it important to understand complement in a more holistic perspective.

Thus given the complexity and importance of complement, developing models of the system within an integrative framework (that can include other biochemical networks) is crucial to understanding its role in human disease. Traditionally, complement models have been formulated as linear or non-linear Ordinary Differential Equation (ODE) systems. Hirayama  et al. used a system of linear ODEs to model the classical pathway of complement \cite{hirayama1996linear}. Korotaevskiy and co-workers built a theoretical model of complement using a system of non-linear ODEs that included classical, lectin and alternate pathways \cite{korotaevskiy2009non}. However both these studies involve no validation studies with experimental data. Liu et al analyzed the formation of classical and lectin C3 convertases and the regulatory role of C4BP using a system of 45 non-linear ODEs with 85 parameters \cite{liu2011computational}.  Recently, Zewde and co-workers built a detailed mechanistic model of alternative complement activation was built using 107 ODEs and 74 kinetic parameters \cite{zewde2016quantitative}. This model delineated the response of complement on a host cell and a foreign antigen. However, these previous models were largely based upon mechanistic knowledge.  Given the complexity of complement and its interactions with other networks it is unfeasible and computationally expensive to build such large mechanistic models. In addition is much more difficult to experimentally interrogate the response of various complement proteins under different conditions. This also presents with the problem of estimation of a large number of parameters with little or no experimental data. Thus there exists a need to reduce the mechanistic complexity while capturing dynamics of complement accurately.

@@

We tested the predictive power of the reduced order human complement model with validation data sets not used during model training. Six validation data sets were used, three for C3a and C5a respectively at different zymosan concentrations. All kinetic and control parameters were
fixed for the validation simulations. The reduced order model predicted the C3a and C5a time-course profiles at a qualitative level (Fig. \ref{fig-prediction}). For the prediction of the $C3a$ at the three different inducer concentrations, our model was able to accurately capture the dynamic behaviors at a qualitative level, including the time the concentration of $C3a$ plateaus. For the prediction for $C5a$, our model was able to capture the behavior at a quantitative level for all three inducer concentrations. However, the our model was not able to capture the concave down curvature for the cases of $0.001~g$ and $0.01~g$ inducer. We believe some of the shortcomings of the prediction may be attributed by our reduced complexity in the modeling of $C4B$ interactions in our system. $C4BP$ interaction was modeled as irreversible mass action kinetics, in reality the binding of $C4BP$ onto other complement components are reversible. Due to this discrepancy in between our model and the experimental results, we believe that the interaction and the reversibility of the $C4BP$ is a key player in creating a threshold where C5a is resistant to minor changes in inducer concentration even though C3a is sensitive.

%This phenomenon may be a part of evolutionary adaptation since C5a is a key factor in the development of many inflammatory diseases such as sepsis, rheumatoid arthritis, and systemic lupus erythemotosis.

@@

Though the role of complement in immune response has been well known since long, there has been a paucity of mathematical models of complement. To our knowledge this is the first model of complement that combines different pathways of initiation and validates the dynamics of downstream proteins like C5a using experimental data. Liu and co-workers modeled formation of C3a through the classical pathway using 45 non-linear ODEs. 
The hybrid modeling framework, however allowed us to model the lectin mediated formation of C3a using only 5 ODEs.  Though we do not model all the interactions of initiation in detail, especially the cross-talk between lectin and classical pathways like Liu et al. we successfully captured C3a dynamics with respect to different concentrations of lectin initiators. The model was also surprisingly accurate in capturing the quantitative dynamics of C3a and C5a formed from the alternate pathway with only 7 ODEs. The lag phase in the initiation of C5a followed by an accelerated production is also qualitatively similar to predicted C5a time profiles using a theoretical model of complement by Zewde et al with 107 equations. Similarly, our model was able to capture C3a formation from the alternate pathway that showed the same qualitative trends as Zewde et al. We also observe in our model simulation that the quantity of C3a produced in the alternate pathway is nearly 1000 times the quantity of  C5a produced. Though this is in agreement with the experimental data  \cite{morad2015time}, it differs from the theoretical predictions by Zewde et al. who show that C3a is ${10^8}$ times the C5a concentration. It is possible that the experimental system has trace amounts of C5a which leads to this difference. In our mode the time profile of C5a generation from the lectin pathway changes with respect to the quantity of zymosan (the lectin pathway initiator). We see that the lag phase for generation progressively decreases with increasing concentration of initiator. Korotaevskiy et al. show a similar lag phase followed by accelerated production of C5a using a theoretical model of complement for much smaller time scales. Zewde et al. show a similar time profile for C5a generated in the alternate pathway as well. We do observe a similar trend in our model, however given the different time scale in our study we do not observe a very prominent lag phase. We also capture 
Our model also captures the max concentrations of C5a with different initiators and in the alternate pathway. However we do not capture the maximum concentration of C3a with low initiator levels. This is probably due to

Taken together we surprisingly do very well in capturing the dynamics of key complement proteins and observe similar trends as in large mechanistic models.

@@

\subsection*{Global sensitivity analysis of model performance}
We conducted a global sensitivity analysis, using the variance-based method of Sobol, to estimate which parameters controlled the performance of the reduced order model \citep{zhang2015sobol}. We computed the total sensitivity index of each parameter relative to four performance objectives, each objective was based on the sum of squared errors between model and experimental data for C3a alternate, C5a alternate, C3a lectin, and C5a lectin simulations. We established the sampling bounds for each parameter from the minimum and maximum value of that parameter in the parameter set ensemble.
We used the sampling method of Saltelli \textit{et al.} \citep{saltelli2010variance} to compute a family of $N\left(2d+2\right)$ parameter sets which obeyed our parameter ranges,
where $N$ was the number of trials, and $d$ was the number of parameters in the model. In our case, $N$ = 200 and $d$ = 28, so the total sensitivity indices were computed from
11,600 model evaluations. The variance-based sensitivity analysis was conducted using the SALib module encoded in the Python programming language \citep{SALIB}.

@@

\begin{figure}[h]
\centering
\includegraphics[width=1.0\textwidth]{./figs/Figure4_Sensitivity_Analysis_v2.pdf}
\caption{Sobol's sensitivity analysis of the reduced order complement model with respect to the modeling parameters.  Sensitivity analysis was conducted on the four cases we used to train our model: (A) C3a at 0 zymosan, (B) C5a 0 zymosan, (C) C3a 1 g zymosan, and (D) C5a $1$ g zymosan. The bars denote total sensitivity index which includes local contribution of each parameter and global sensitivity of significant pairwise interactions. The error bars are the 95 percent confidence interval. $k$ represents association rate, $km$ denote Michaelis-Menten saturation constants, and $alpha$ and $n$ refers to the exponentials of the control functions.}\label{fig-SA}
\end{figure}
